---
title: "Testing for scale-dependent phylogenetic signal in intraspecific variation in intertegular distance across bee communities in Montana, USA"
output: pdf_document
date: "2024-10-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Species interactions vary across space and time due to both species turnover and to a phenomenon known as interaction rewiring. Interaction rewiring may be driven by temporal or spatial changes in phenology, abundance, or species traits (both intra- and interspecific). The impacts of interspecific trait matching, phenology, and abundance on ecological netowrks have been repeatedly demonstrated: changes in flowering time or tongue morphology have clear impacts on plant-pollinator interactions, for example. However, the role of intraspecific trait differences in shaping interactions—and their relative importance—remain poorly known. 

Bees are a species-rich (n>20,000) clade of Hymenopterans with a cosmopolitan distribution. Ecologically important pollinators, intraspecific variation in body size is common in some taxa. Because body size is highly correlated with tongue length and flying ability, it is plausible it mediates foraging choices and thus may explain a portion of plant-pollinator interaction rewiring across space. 

## Methods

- Bees and flowering plants were sampled when observed interacting across 4 years and 142 sites in Montana, USA. Sites were arranged into spatially nested design: 9 sites comprised a block and 8 blocks comprised a locality; there were two localities in the study. 
- Bees were associated with their foraging substrate, collected and pinned for identification. A proxy for body size, intertegular distance (ITD)—the distance between tegulae (wing bases) on the—was measured in each female individual. (As 75% of collected individuals were female, male bees were excluded.)

## Questions

Our study aims to answer the following questions:

1) How much phylogenetic signal is there in the magnitude of intraspecific variation in female bee body size?
2) At what spatial scale(s) does phylogenetic signal emerge?
3) Do bee species present at a greater number of sites have more intraspecific variation in female bee body size?
4) Is intraspecific variation in female bee body size predicted by interaction partners after controlling for phylogeny?
5) Does the strength of the relationship between intraspecific variation in female bee body size vary across scales?

## Analysis

We'll begin by loading libraries for data manipulation, visualization, and Bayesian phylogenetic mixed models:

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(reshape2)
library(brms)
library(ape)
library(cowplot)
library(ggtree)
library(phytools)
```

Let's also write a function to calculate Bao's estimator of variation ($CV_4$):

```{r, message=FALSE, warning=FALSE}
bao <- function(data){
  N=length(data)
  y_bar=mean(data)
  s2_hat=var(data)
  cv_2=s2_hat/y_bar^2
  cv_1=sqrt(cv_2)
  gamma_1=sum(((data-y_bar)/s2_hat^0.5)^3)/N
  gamma_2=sum(((data-y_bar)/s2_hat^0.5)^4)/N
  bias=cv_2^(3/2)/N*(3*cv_2^0.5-2*gamma_1)
  bias2=cv_1^3/N-cv_1/4/N-cv_1^2*gamma_1/2/N-cv_1*gamma_2/8/N
  cv4=cv_1-bias2
  return(cv4)
}

```

Now we load our data:

```{r, message=FALSE, warning=FALSE}
bee_data <- read_csv("~/Dropbox/bee_phylo_itv/bee_females_for_ethan.csv") # bee trait data
bee_phylo <- read.tree("~/Dropbox/bee_phylo_itv/BEE_prunedtree.nwk") # pruned newick-format phlogeny of bee species present in MT
```

Let's look at the trait data first: 

```{r, message=FALSE, warning=FALSE}
bee_data
```

Important columns include `TRANSECT_COMBO`, `LOCATION`, `block_2`, `IT_microns`, `GENUS_SPECIES`, and `FLOWER_SPECIES`. We can summarize these data to get an idea of their size and diversity. First, the number of bee species in the dataset:

```{r, message=FALSE, warning=FALSE}
bee_data$GENUS_SPECIES %>% unique() %>% length()
```

...the number of plant species: 

```{r, message=FALSE, warning=FALSE}
bee_data$FLOWER_SPECIES %>% unique() %>% length()
```

...and the total number of observations: 

```{r, message=FALSE, warning=FALSE}
bee_data %>% nrow()
```

We may need to filter down these data to only include female bees. Let's check to see what levels are present in the "SEX" column:

```{r, message=FALSE, warning=FALSE}
bee_data$SEX %>% unique()
```

None, apparently—we must be working with a pre-filtered CSV. In that case, let's go ahead and calculate Bao's CV for all species with a sample size of n $\geq$ 20: 

```{r, message=FALSE, warning=FALSE}
set.seed(1) # set seed for reproducibility
bao_values <- bee_data %>%
  group_by(GENUS_SPECIES) %>%  
  mutate(total_observations = n()) %>% 
  filter(n() >= 20) %>%                        # Filter groups with 20 or more observations
  sample_n(20) %>%                             # Randomly sample 20 rows from the filtered data
  mutate(bao_cv = bao(IT_microns), 
    num_sites = n_distinct(TRANSECT_COMBO)) %>%
  select(GENUS_SPECIES, bao_cv, num_sites, total_observations) %>%
  distinct()
bao_values
```

How are these data distributed?

```{r, message=FALSE, warning=FALSE}
p1 <- ggplot(bao_values, aes(x=log(bao_cv))) + 
  theme_bw() +
  geom_histogram(bins=40)
p1
```

Next, let's prepare our phylogenetic data. We need to create a consensus phylogeny that retains branch lengths (by averaging), drop tips that aren't present in the 59 species in our summary dataset, cand alculate and visualize a correlation matrix given a Brownian motion model of trait evolution (NOT a covariance matrix; see dicussion here: https://discourse.mc-stan.org/t/covariance-matrix-phylogenetic-models/20477/4). 

```{r, message=FALSE, warning=FALSE}
bee_phylo_consensus <- consensus.edges(bee_phylo) 
bee_phylo_consensus <- keep.tip(bee_phylo_consensus, unique(bao_values$GENUS_SPECIES))
A <- ape::vcv.phylo(bee_phylo_consensus, corr = TRUE)
melted_A <- melt(A)
head(melted_A)
cov <- ggplot(data = melted_A, aes(x=Var1, y=Var2, fill=value)) + 
  geom_tile() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), axis.title = element_blank())
cov
```

Here, areas of light blue show high correlation values (e.g., are pairwsie comparisons of members of the same clade). 

## 1) How much phylogenetic signal is there in the magnitude of intraspecific variation in female bee body size?

To test for phylogenetic signal in our ITV, we'll build a simple Bayesian linear model to predict Bao's CV from the number of sites where a given species is found. In out subsampled data, there *may* be a weak association between these variables, as well as the total number of observations of a species (before subsetting) and its Bao CV value (calculated after subsetting): 

```{r, message=FALSE, warning=FALSE}
p2 <- ggplot(bao_values, aes(x=num_sites, y=log(bao_cv), )) + 
  theme_bw() +
  geom_point(pch=21)
p3 <- ggplot(bao_values, aes(x=total_observations, y=log(bao_cv), )) + 
  theme_bw() +
  geom_point(pch=21) +
  xlim(0,100) # note trimmed axis to ignore outliers

plot_grid(p2, p3)
```

In a Bayesian statistics, a linear model will tell us the posterior probability of its parameters (i.e., the effect size of a predictor) given the data. Ignoring the marginal likelihood (or "probability of data"), this is what that looks like:    

$$
P(\text{model parameters}|\text{data}) \propto P(\text{data}|\text{model parameters})*P(\text{model parameters})
$$

It's useful to formally define our model. We start with the likelihood, or the probability of a particular parameter value given the observed data (i.e., the first hald of the right-hand side of the equation). It is a little tricky to think about the most appropriate probability distribution for Bao's CV. The data-generating process—local adaptation to different niches—might lead to gamma-distributed body sizes (in this case, right-skewed data). Gamma distributed data seem tricky to deal with in linear models. By log-transforming Bao's CV for intertegular distance we should be able to use a Gaussian distribution, which will make calculating phylgoenetic signal easier. The likehood thus tells us Bao's CV is stochastically related to a normal distribution with parameters $\mu_i$ (the mean) and $\sigma$ (variance). In this case, $\mu_i$ is itself *deterministically* related to the parameter $\beta$ ("number of sites"), with an overall intercept ($\alpha$) plus a set of phylogenetically correlated intercepts ($\alpha_j$): 

$$
\text{log(Bao's CV)} \sim \text{Normal}(\mu_i, \sigma)
$$
$$
\mu_i = \alpha + \alpha_j + \beta_{\text{abundance},i}
$$

We then need to define our priors, or the distribution of sensible values for model parameters. We know that $\alpha_j$ should be drawn from a normal distribution with mean $\alpha$ and correlation structure $\sigma_A$:

$$
\alpha_j \sim \text{Normal}(\alpha, \sigma_A)
$$

At this point it's probably easiest to get the rest of the priors we need from the `get_prior` function from `brms`:

```{r, message=FALSE, warning=FALSE}
priors <- get_prior(bao_cv ~ 1 + num_sites + (1|gr(GENUS_SPECIES, cov = A)),
                    data = bao_values, data2 = list(A = A), family=gaussian())   
prior
```

This gives us a flat prior for the effect of num_sites on the outcome variable, e.g. a uniform distribution over all reals: 

$$
\beta_{\text{abundance},i} \sim \text{Uniform}(-\infty,\infty)
$$

It also suggests that we use a standard Student's T distribution for $\sigma$, the standard deviation of the residuals:

$$
\sigma \sim \text{Student}(3, 0, 2.5)
$$

Now, we'll run our model, using an MCMC chain with the default number of steps: 

```{r, message=FALSE, warning=FALSE, results=FALSE}
model_1 <- brm(
  log(bao_cv) ~ num_sites + (1|gr(GENUS_SPECIES, cov = A)),
  data = bao_values,
  family = gaussian(), 
  data2 = list(A = A),
  prior = priors
)
```
(Output curtailed to keep this document a manageable size/)

Let's summarize the results:

```{r, message=FALSE, warning=FALSE}
summary(model_1)
```

This suggests there is a positive effect of abundance on Bao's CV of intertegular distance—but that it is not credible at the 95% level. It also suggests that phylogenetically correlated intercepts explain a lot of the variation in Bao's CV we see here. We can formally calculate Pagel's $\lambda$—the proportion of variance in the model attributable to phylogeny—using its residuals. We can also perform a one-sided hypothesis test to evaluate the evidence that $\lambda > 0$. To do so, we calculate the ratio of the posterior probabilty that $\lambda>0$ compared with the posterior probability that $\lambda=0$: 

```{r,  message=FALSE, warning=FALSE}
hyp <- "sd_GENUS_SPECIES__Intercept^2 / (sd_GENUS_SPECIES__Intercept^2 + sigma^2) = 0"
(hyp <- hypothesis(model_1, hyp, class = NULL))
plot(hyp)
```

This appears to be greater than Laura's estimates from `phytools`. The obvious next steps would be to see whether signal declines using small subsets of the data (e.g., ask whether $\lambda$ only emerges at broader spatial scales). But we should chat first, I think!

